ProteInfer: deep networks for protein functional inference

### 2021-10-06 latex/draft3.tex
Updates

Beyond functional annotation, deep learning has enabled significant advances in protein structure prediction \cite{alquraishi2019end, senior2020improved, Yang1496, yang2020improved, du2020energy, rao2021msa}, predicting the functional effects of mutations \cite{riesselman2018deep,rives2019biological,rao2019evaluating,frazer2020large}, and protein design \cite{yang2019machine, mazurenko2019machine, biswas2020low, madani2020progen,elnaggar2020prottrans, anishchenko2020novo, bryant2021deep}. A key departure from traditional approaches is that researchers have started to incorporate vast amounts of raw, uncurated sequence data into model training, an approach which also shows promise for functional prediction \citep{Brandes2021.05.24.445464}.

@@

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are alternative methods for splitting \citep{Brandes2021.05.24.445464,zhou2019cafa, gillis}, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. Such a split is not available for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives \citep{warwick2020benchmarking}.

>>>>

Beyond functional annotation, deep learning has enabled significant advances in protein structure prediction \cite{alquraishi2019end, senior2020improved, Yang1496, yang2020improved, du2020energy, rao2021msa}, predicting the functional effects of mutations \cite{riesselman2018deep,rives2019biological,rao2019evaluating,frazer2020large}, and protein design \cite{yang2019machine, mazurenko2019machine, biswas2020low, madani2020progen,elnaggar2020prottrans, anishchenko2020novo, bryant2021deep}. A key departure from traditional approaches is that researchers have started to incorporate vast amounts of raw, uncurated sequence data into model training, an approach which also shows promise for functional prediction \citep{Brandes2021}.

@@

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are alternative methods for splitting \citep{Brandes2021,zhou2019cafa, gillis}, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. Such a split is not available for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives \citep{warwick2020benchmarking}.
